Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2020-04-06 treestats_paper.tex
correction from Leo

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at
scale~\citep{kelleher2019inferring,speidel2019method}.
In Figure~\ref{fig:sweep_duality}, we showed that Branch and Site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct (which inference methods can only hope to approximate).
To evaluate the correspondence between Site and Branch statistics
in trees inferred from real data,
we calculated statistics for chromosome 20
of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
by Relate~\citep{speidel2019method}. \revpoint{2}{9}
(Although Relate estimates a succession of essentially independent marginal trees rather
than a succinct tree sequence,
the output can be converted to a tree sequence, and since the statistics considered
here are ``single site'' the distinction is not important.)
Specifically, we calculated diversity using the Site statistic
described in Example~\ref{ex:site_diversity},
and compared this to the dual Branch statistic (Example~\ref{ex:branch_diversity})
in 1Mb windows in each of the five continental groupings.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
As shown in Figure~\ref{fig:chr20}, the ratio of Site and Branch diversity
is relatively constant, hovering around $2.5 \times 10^{-8}$.
This is somewhat higher than
typical estimates for the average genome-wide human mutation rate~\citep[e.g., $1.45\times 10^{-8}$,][]{narasimhan2017estimating},
which suggests a mis-calibration in inferred tree times.
These trees were inferred using Relate with the Ne fixed to 20,000;
if trees are instead jointly inferred in Relate along with an Ne, with mutation rate fixed,
then indeed the ratio of Site to Branch diversity averages to the (externally specified) mutation rate
(\autoref{sfig:chr20_GBR}).
On the one hand, some amount of constancy of this ratio is expected, since
Relate estimates branch lengths in part assuming that mutations fall at a constant rate through time.
(But, note that \citeauthor{speidel2019method} also showed signal of changing mutation rates,
pointing the way towards a more general method.)
On the other hand,
since estimated node ages are shared across long regions of the genome,
a tight agreement may not be possible with poorly inferred trees.
So, the relative constancy of the ratio of Site to Branch diversity
suggests that Relate is doing a good job at inferring trees,
but what to make of the twofold variation in this ratio?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.

>>>>

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at
scale~\citep{kelleher2019inferring,speidel2019method}.
In Figure~\ref{fig:sweep_duality}, we showed that Branch and Site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct (which inference methods can only hope to approximate).
To evaluate the correspondence between Site and Branch statistics
in trees inferred from real data,
we calculated statistics for chromosome 20
of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
by Relate~\citep{speidel2019method}. \revpoint{2}{9}
(Although Relate estimates a succession of essentially independent marginal trees rather
than a succinct tree sequence,
the output can be converted to a tree sequence, and since the statistics considered
here are ``single site'' the distinction is not important.)
Specifically, we calculated diversity using the Site statistic
described in Example~\ref{ex:site_diversity},
and compared this to the dual Branch statistic (Example~\ref{ex:branch_diversity})
in 1Mb windows in each of the five continental groupings.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
As shown in Figure~\ref{fig:chr20}, the ratio of Site and Branch diversity
is relatively constant, hovering around $2.5 \times 10^{-8}$.
This is somewhat higher than
typical estimates for the average genome-wide human mutation rate~\citep[e.g., $1.45\times 10^{-8}$,][]{narasimhan2017estimating},
which suggests a mis-calibration in inferred tree times.
These trees were inferred using Relate with the Ne fixed to 15,000;
if trees are instead jointly inferred in Relate along with an Ne, with mutation rate fixed,
then indeed the ratio of Site to Branch diversity averages to the (externally specified) mutation rate
(\autoref{sfig:chr20_GBR}).
On the one hand, some amount of constancy of this ratio is expected, since
Relate estimates branch lengths in part assuming that mutations fall at a constant rate through time.
(But, note that \citeauthor{speidel2019method} also showed signal of changing mutation rates,
pointing the way towards a more general method.)
On the other hand,
since estimated node ages are shared across long regions of the genome,
a tight agreement may not be possible with poorly inferred trees.
So, the relative constancy of the ratio of Site to Branch diversity
suggests that Relate is doing a good job at inferring trees,
but what to make of the twofold variation in this ratio?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.
